biclaR: Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools

### 2023-12-21 paper/PaperTRA/PaperTRA_rev1.Rmd
setup PaperTRA_rev1 redux 6p
as requested by reviewers
reduce do 6+1

abstract: | 
  A high proportion of car trips can be replaced by a combination of public transit and cycling for the first-and-last mile. This paper estimates the potential for cycling combined with public transit (PT) as a substitute for car trips in the Lisbon metropolitan area and assesses its socio-environmental impacts using open data and open source tools.
  A decision support tool that facilitates the design and development of a metropolitan cycling network was developed (_biclaR_). The social and environmental impacts were assessed using the _HEAT for Cycling_ and the _HEAT as a Service_ tools. The impacts of shifting car trips to PT were also estimated and monetized.
  The results indicate that 10% of trips could be made by bicycle + PT combination. Shifting to cycling for the first-and-last mile stages can reduce annual CO~2~eq emissions from 6,000 tons/day, with benefits over 10 years of at least €230 million. For the PT leg, the transfer from car avoids of up to 20,500 tons of CO~2~eq emissions per year.
  This evidence can support policymakers to prioritize interventions that reduce the reliance on private motor vehicles.
keywords: 
  - Active transport
  - Intermodality
  - First and last mile
  - Health economic assessment
  - Environmental impacts
  - Open data and methods
bibliography: bibliography.bib
# Use csl when using default citation_package for Pandoc citeproc
csl: https://www.zotero.org/styles/springer-lecture-notes-in-computer-science
# cls : svproc.csl
# citations with arabic numbers is prefered when using natbib
natbiboptions: numbers

@@

<!-- **full paper**: 4-6 pages in length (typically up to 3,000 words). we have 2849 words before references, or 2278 words without tables -->

@@

To achieve the cycling targets set by the Portuguese national cycling strategy for 2025 and 2030 (4% and 10%, respectively) [@ENMAC], the Lisbon's Metropolitan Department of Transport introduced *biclaR*[^1], a decision support tool that facilitates the design and development of a metropolitan cycling network [@felix2023].

@@

*biclaR* builds on the Propensity to Cycle Tool[^2] (PCT), a web application and research project funded by the UK's Department for Transport in 2015 which launched nationally in 2017 as part of the government's Cycling and Walking Investment Strategy.
The PCT initially used only origin-destination data for commuting trips as the basis of estimates of cycling potential at zone, route and route network levels [@lovelace2017].
The PCT has been extended to include cycling potential for travel to school in England [@goodman2019] and other trip types in other countries.[^3]
However, to the best of our knowledge, this is the first time that the method has been integrated with public transport data using multi-modal routing to estimate the potential and benefits of multi-stage cycling and PT trips.

@@

[^3]:  See [npt.scot](https://www.npt.scot) and [cruse.bike](https://cruse.bike) for examples of the PCT in Scotland and Ireland that include estimates of cycling for other purposes.

@@

We used a method for disaggregating the origins and destinations of trips between the centroids of two districts (same as "parish") to ensure that a district is not solely characterized by a single point of origin and destination for its trips.
Aggregating all trips into centroids renders the exercise less realistic, as it excludes a significant portion of short-distance trips, a prevalent characteristic of active mode travel [@Lovelace2022Jittering].
The OD Jittering method breaks down a single point (i.e., the centroid of an area) into multiple random points on the existing and neighboring road network, using OpenStreetMap as a reference.
This method then distributes the volume of trips within the district among the randomly generated origin-destination pairs.

Using the [`odjitter` R package](https://github.com/dabreegster/odjitter), we employed a maximum disaggregation level of 100 trips per O-D pair for this project.
Figure \ref{fig:jitter} illustrates the contrast between trip representation through the traditional method, which connects a single desire line between each district, and the presentation achieved through the randomization and disaggregation of trips between districts, specifically for the Lisbon metropolitan area.

```{r jitter, fig.align='center', fig.cap="Representation of OD pairs in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).", out.width="100%"}

@@

Although this method provides a more realistic representation of the trips undertaken compared to the traditional approach, it does not fully align with the actual O-D pairs of trips, which remain unknown due to data privacy regulations.

@@

The mobility survey collects the origin and destination of trips but does not include the respective routes.
Modeling the realistic cycling + PT routes between OD pairs depends on assumptions regarding the characteristics of the cycling and road networks and the location of public transport interfaces.
Other constraints regarding the behavior of potential cyclists determine the routing results.
For example, such restrictions can favor low speed, low traffic streets, more direct routes, and less steep paths, among others, suitable for cycling.

@@

The selected route choice algorithm was the [`r5r` R package](https://ipeagit.github.io/r5r/) [@r5r], which allows for great flexibility in configuring estimated route types, and which proven to provide most accurate route networks for the city of Lisbon [@Lovelace2022exploring].
`r5r` can calculate multi-modal routes using PT combined with other modes.
It enables the identification of the most direct or safest cycling routes, using the Level of Traffic Stress[^4] (LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest (e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g., routes shared with motorized traffic).
The routes were estimated for the base scenario for both types of networks: *direct* and *safe*, using LTS 4 and LTS 3, respectively.
Different routing profiles enable decision-makers to plan for different bicycle user typologies and/or for different city cycling maturity levels [@felix2017]. <!-- move this to conclusions? -->

@@

The `r5r` model used the OpenStreetMap road network and the GTFS metropolitan data aggregated and validated.
This information is crucial for an accurate PT trip and route estimation.
A digital elevation model, from the European Space Agency's COPERNICUS mission, was used to include street gradient information, as a weight in cycling routing. <!--  with a 25 m spatial resolution, -->
The cycling potential trips for the two national strategic targets (4% and 10%) were estimated from the 2017 cycling and car trips (both as a driver and as a passenger), the baseline scenario.

The routes were then overlaid and aggregated by segments, using [`stplanr overline()` R function](https://docs.ropensci.org/stplanr/reference/overline.html).
<!-- For example, the estimated volumes of bicycle trips and car transfers were aggregated, and the average car speed and quietness level, weighted by the distance of the segments, were calculated. -->

@@

The intermodality scenario considers trips combining PT and cycling for the first and last legs.
In a conservative approach, we have restricted our analysis to the first and last legs with a combined length of up to 5 km (for instance: 1 km from origin to interface A plus 4 km from interface B to destination) or up to 25 minutes on bike.
Furthermore, we have imposed restrictions on PT usage to include only trips with no PT transfers, and up to 2 hours (120 min).
Additionally, we have only included PT modes that can easily accommodate bicycles, such as trains, ferries, trams, and inter-municipal bus lines equipped with bike racks (Figure \ref{fig:map1}).

```{r map1, out.width="60%", fig.cap="Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area", fig.align='center'}
knitr::include_graphics("img/map1.png", error = FALSE)
```

Figure \ref{fig:map2} illustrates the resulting bicycle routes to access the main PT interfaces in the LMA.

```{r map2, out.width="80%", fig.cap="Bike routes with highest potential to serve as first and last leg when replacing cycling and PT from car trips (screenshot of the interactive online tool).", fig.align='center'}
knitr::include_graphics("img/map2.png", error = FALSE)
```

@@

For the cycling legs of the journey (first and last legs), socio-environmental impacts were estimated, using the HEAT for Cycling tool v5.0 [@HEAT] from the World Health Organization, and the [`HEATaaS` R package](https://github.com/HEAT-WHO/HEAT_heatr_api)[^5].
The HEAT tool provided estimates on the shifting from car to cycling for a short term time horizon (i.e., one year) and the long term (i.e., ten years).
We considered two dimensions: *social* — including the physical activity of cyclists, air pollution exposure, and road casualties; *environmental* — including CO~2~eq emissions and other pollutants.

@@

To estimate the car emissions, we used the EMEP/EEA's COPERT software v5 methods and reference values [@COPERT] for a Tier 3 detail level. 
We used a family-size vehicle, EURO standard, and gasoline or diesel fuel.
All trips were considered to be made under urban conditions and at an average speed of 15 km/h during rush hour periods.
Since the average distance traveled per trip influences the overconsumption and emissions from cold-start engine operation, we estimated energy and emission factors for different ranges of trips at 500-meter intervals.
In addition, we assumed an occupancy rate of 1.6 passengers *per* car [@IMOB].

Regarding PT, we considered the emission factor values reported in the environmental and sustainability reports of the PT operators in the LMA [@Carris2019s; @Metro2019s; @CP2019s; @Transtejo2014].

Emissions were estimated for the following atmospheric pollutants: CO, PM10, NOx, and VOC; and for the main green house gases: CO~2~, CH~4~, and N~2~O, converted in CO~2~eq.
In particular, for the urban train and tram -- with 100% electric traction -- only CO~2~eq emissions were considered (resulting from the production of electricity, considering a "well-to-tank" approach), since the other pollutants are not emitted locally.

The conversion of avoided emissions into avoided welfare loss and respective monetary valuation was based on the EU Guide to Cost-benefit Analysis [@EuropeanCommission2014] and the best up-to-date reference values for the various gases [@EuropeanCommission2014; @bickel2006; @UNITE].
The social impacts are in avoided premature mortality. This result is finally monetized using the *Statistical Value of Life* for Portugal [@ANSR2021].
We updated all the monetary reference values of the literature based on the annual inflation rate in Portugal for 2022[^6], and our 10-years estimations assumed a discount rate of 5% and inflation of 3%.

[^6]: See [Statistics Portugal tool for inflation rate estimates between years](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ipc).

@@

```{r summary1}
summary1 = readRDS("load/summary1.Rds")

summary1 = summary1 %>%
  mutate(LTS = recode(LTS, `3` = "safe", `4` = "direct")) |> 
  mutate(Estrategia = paste0(Estrategia*100,"%")) |> 
  mutate(TotalTrips = TotalTrips/2,
         Bikebase = Bikebase/2,
         Bikepotential = Bikepotential/2) #this is a quick correction while we donr run biclar entirely again. The error was found and it was adding up the 2 bike legs, instead of making the mean. The SE estimates remain the same because they are from the bikeshare and carshare percentages, that did not change.

knitr::kable(summary1,
      digits = 0,
      format.args = list(big.mark = " "),
      col.names = c("Target", "Routing", "Total trips",
                    "Baseline Cycling + PT", "Potencial Cycling + PT",
                    "Avoided CO2eq (ton/yr)", "SE Benefits for 10 years (thousand €)"),
      caption = "\\label{summary1}Summary of the cycling potencial of intermodality scenario and its socio-environmental benefits for the cycling legs.",
      format="latex",
      booktabs = TRUE) |> 
  kableExtra::column_spec(column = c(4:7), width = "6em")

@@

Table \ref{tab:summary21} shows the potential trips by PT mode to replace the second leg of the journey, in combination with cycling.
Trains offer the greatest potential for substitution (88%).
When comparing the existing PT interfaces (Figure \ref{fig:map1}) with the bike routes with highest potential to serve as first and last legs (Figure \ref{fig:map2}) it becomes clear that the Train interfaces are the ones that have the highest potential to attract car-to-PT substituting trips, if their accessibility by bicycle is improved to be safe.

```{r summary21}
summary2 = readRDS("load/summary2.Rds")

@@

Table \ref{tab:summary22} presents the results of the avoided emissions and its monetization for the second leg of the journey, by replacing car trips with potential TP trips.
<!-- We estimate a reduction of 8,702 tons of CO2eq emissions by substituting motorized trips relying on fossil fuels and electricity. These estimations used an electricity production life cycle approach. -->
<!-- this table could be split into Social benefits and Environmental benefits -->
Regarding the PT segment, the shift from private car usage would lead to the mitigation of CO~2~ equivalent emissions to 8,500 to 20,800 tons annually, valued in €1.4 million to €3.5 million yearly.

```{r summary22}
summary22 = readRDS("load/TPtable_aml.Rds")
summary22 = summary22 %>%
  mutate(LTS = recode(Percursos, "seguro" = "safe", "direto" = "direct")) |> 
  mutate(Value = money_emissions_eur/1000) |> 
  select(Meta_ENMAC, LTS, CO2eq_TP, CO_PT, PM10_PT, NOx_PT, VOC_PT, Value) |> 
  arrange(Meta_ENMAC)

knitr::kable(summary22,
      digits = c(0,0,0,0,1,0,1,0),
      format.args = list(big.mark = " "),
      col.names = c("Target", "Routing", "CO2eq",
                    "CO", "PM10", "NOx", "VOC", "Value (k€)"),
      caption = "\\label{summary22}Summary of the avoided emmissions (ton/year) and corresponding monetization (thousand €) by replacing car trips with PT, in the second leg.",
      format="latex",
      booktabs = TRUE) |> 
  kableExtra::column_spec(column = c(3:7), width = "3.5em")
```

@@

```{r summaryall, message=FALSE, warning=FALSE}
summary0 = readRDS("load/HEAT_aml_results.Rds") |> slice_tail(n = 4)
rate = ((1+0.03)^10)/((1+0.05)^10) # inflation rate 3% discount rate 5%

summaryall = summary1 |> 
  mutate(Meta_ENMAC = Estrategia) |>
  left_join(summary22) |> 
  mutate(Code = c(3304, 3404, 3310, 3410)) |> 
  left_join(summary0 |> select(Code, Economic), by = "Code") |> 
  select(Estrategia, LTS, co2, CO2eq_TP, monet, Value, Economic) |> 
  mutate(CO2eq = co2 + CO2eq_TP,
         value1 = Economic/1000 + Value, # this is CORRECT
         value10 = monet + Value*10*rate) |> # this is CORRECT
  select(Estrategia, LTS, CO2eq, value1, value10)

@@

<!-- FOR OTHER PAPER: separate the Socio impacts from the Environmental ones? It shouldn't be ok to sum them -->
Shifting from car to cycling + PT can reduce annual CO~2~eq emissions by 14,000 to 36,000 tons per year, 
and the 10-year socio-environmental benefits account for €235 million to €620 million, depending on the cycling targets.

The social impacts represent 98% of the socio-environmental benefits (in value) from replacing car trips to bicycle in first-and-last legs. For the PT segment, we did not estimate the social impacts from substituting car trips, although its health benefits would not be as high as shifting to cycling.<!-- We should estimate the social impacts from replacing Car to PT in the second leg, for the other paper --> <!--Overall there was very little evidence available on mode switch to public transport. -->

The emissions of CO~2~eq that are avoided during both the initial and final journey segments account for about 75% of the emissions avoided during the PT segment. This finding, while expected -- due the zero cycling emissions, should not be overlooked when promoting the PT use.
Improving the safe accessibility to PT interfaces to cyclists and providing bicycle-friendly amenities such as parking facilities can potentially lead to a higher reduction in CO~2~eq emissions, compared to a scenario where individuals shift from car travel to car + PT combination.

Our findings suggest that cycling and PT *in combination* could viably replace 10% of current LMA trips, with an additional 6% of PT journeys prone to further substitution. 
<!-- repeated? not correct -->

>>>>

abstract: | 
 This paper estimates the potential for cycling combined with public transit (PT) as a substitute for car trips in the Lisbon metropolitan area and assesses its socio-environmental impacts using open data and open source tools.
  A decision support tool that facilitates the design and development of a metropolitan cycling network was developed (_biclaR_). The social and environmental impacts were assessed using the _HEAT for Cycling_ and the _HEAT as a Service_ tools. The impacts of shifting car trips to PT were also estimated and monetized.
  The results indicate that 10% of trips could be made by bicycle + PT combination. Shifting to cycling for the first-and-last mile stages can reduce annual CO~2~eq emissions from 3,000 tons/day, with benefits over 10 years of at least €125 million. For the PT leg, the transfer from car avoids of up to 20,500 tons of CO~2~eq emissions per year.
  This evidence can support policymakers to prioritize interventions that reduce the reliance on private motor vehicles.
keywords: 
  - Active transport
  - Intermodality
  - First and last mile
  - Health economic assessment
  - Environmental impacts
  - Open methods
bibliography: bibliography.bib
# Use csl when using default citation_package for Pandoc citeproc
csl: https://www.zotero.org/styles/springer-lecture-notes-in-computer-science
# cls : svproc.csl
# citations with arabic numbers is prefered when using natbib
natbiboptions: numbers

@@

<!-- **full paper**: 4-6 pages in length (typically up to 3,000 words). we have 2849 words before references, or 2278 words without tables -->
<!-- REDUCE TO MAX 6pages -->

@@

To achieve the cycling targets set by the Portuguese national cycling strategy for 2025 and 2030 (4% and 10%, respectively), the Lisbon's Metropolitan Department of Transport introduced *biclaR*[^1], a decision support tool that facilitates the design and development of a metropolitan cycling network [@felix2023].
<!-- [@ENMAC] -->

@@

*biclaR* builds on the Propensity to Cycle Tool[^2] (PCT), a web application and research project funded by the UK's Department for Transport in 2015 which launched nationally in 2017 as part of the government's Cycling and Walking Investment Strategy.
The PCT initially used only origin-destination data for commuting trips as the basis of estimates of cycling potential at zone, route and route network levels [@lovelace2017].
<!-- The PCT has been extended to include cycling potential for travel to school in England [@goodman2019] and other trip types in other countries.[^3] -->
However, to the best of our knowledge, this is the first time that the method has been integrated with public transport data using multi-modal routing to estimate the potential and benefits of multi-stage cycling and PT trips.

@@

<!-- [^3]:  See [npt.scot](https://www.npt.scot) and [cruse.bike](https://cruse.bike) for examples of the PCT in Scotland and Ireland that include estimates of cycling for other purposes. -->

@@

We used a method for disaggregating the origins and destinations of trips between the centroids of two districts (same as "parish") to ensure that a district is not solely characterized by a single point of origin and destination for its trips.
<!-- Aggregating all trips into centroids renders the exercise less realistic, as it excludes a significant portion of short-distance trips, a prevalent characteristic of active mode travel. -->
The OD Jittering method breaks down a single point (i.e., the centroid of an area) into multiple random points on the existing and neighboring road network, using OpenStreetMap as a reference [@Lovelace2022Jittering].
<!-- This method then distributes the volume of trips within the district among the randomly generated origin-destination pairs. -->
Using the [`odjitter` R package](https://github.com/dabreegster/odjitter), we employed a maximum disaggregation level of 100 trips per O-D pair for this project.
<!-- Figure \ref{fig:jitter} illustrates the contrast between trip representation through the traditional method, which connects a single desire line between each district, and the presentation achieved through the randomization and disaggregation of trips between districts, specifically for the Lisbon metropolitan area. -->
Although this method provides a more realistic representation of the trips undertaken compared to the traditional approach, it does not fully align with the actual O-D pairs of trips, which remain unknown due to data privacy regulations.

```{r jitter, fig.align='center', fig.cap="Representation of OD pairs in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).", out.width="100%", eval=FALSE}

@@

@@

The mobility survey collects the origin and destination of trips but does not include the respective routes.
Modeling the realistic cycling + PT routes between OD pairs depends on assumptions regarding the characteristics of the cycling and road networks and the location of public transport interfaces.
<!-- Other constraints regarding the behavior of potential cyclists determine the routing results. -->
<!-- For example, such restrictions can favor low speed, low traffic streets, more direct routes, and less steep paths, among others, suitable for cycling. -->

@@

The selected route choice algorithm was the [`r5r` R package](https://ipeagit.github.io/r5r/) [@r5r], which allows for great flexibility in configuring estimated route types.
<!-- and which proven to provide most accurate route networks for the city of Lisbon [@Lovelace2022exploring]. -->
`r5r` can calculate multi-modal routes using PT combined with other modes.
It enables the identification of the most direct or safest cycling routes, using the Level of Traffic Stress[^4] (LTS) scale.
<!-- ranging from 1 to 4, where 1 corresponds to the quietest (e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g., routes shared with motorized traffic). -->
The routes were estimated for the base scenario for both types of networks: *direct* and *safe*, using LTS 4 and LTS 3, respectively.
<!-- Different routing profiles enable decision-makers to plan for different bicycle user typologies and/or for different city cycling maturity levels [@felix2017]. <!-- move this to conclusions? -->

@@

The `r5r` model used the OpenStreetMap road network and the GTFS metropolitan data aggregated and validated.
This information is crucial for an accurate PT trip and route estimation.
<!-- A digital elevation model, from the European Space Agency's COPERNICUS mission, was used to include street gradient information, as a weight in cycling routing. <!--  with a 25 m spatial resolution, -->
<!-- The cycling potential trips for the two national strategic targets (4% and 10%) were estimated from the 2017 cycling and car trips (both as a driver and as a passenger), the baseline scenario. -->

<!-- The routes were then overlaid and aggregated by segments, using [`stplanr overline()` R function](https://docs.ropensci.org/stplanr/reference/overline.html). -->
<!-- For example, the estimated volumes of bicycle trips and car transfers were aggregated, and the average car speed and quietness level, weighted by the distance of the segments, were calculated. -->

@@

The intermodality scenario considers trips combining PT and cycling for the first and last legs.
In a conservative approach, we have restricted our analysis to the first and last legs with a combined length of up to 5 km (for instance: 1 km from origin to interface _A_ plus 4 km from interface _B_ to destination) or up to 25 minutes on bike.
Furthermore, we have imposed restrictions on PT usage to include only trips with no PT transfers, and up to 2 hours (120 min).
Additionally, we have only included PT modes that can easily accommodate bicycles, such as trains, ferries, trams, and inter-municipal bus lines equipped with bike racks (Figure \ref{fig:map1}).
Figure \ref{fig:map2} illustrates the resulting bicycle routes to access the main PT interfaces in the LMA.

```{r map1, out.width="60%", fig.cap="Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area", fig.align='center', fig.pos="h"}
knitr::include_graphics("img/map1.png", error = FALSE)
```

```{r map2, out.width="80%", fig.cap="Bike routes with highest potential to serve as first and last leg when replacing cycling and PT from car trips (screenshot of the interactive online tool).", fig.align='center', fig.pos="h"}
knitr::include_graphics("img/map2.png", error = FALSE)
```

@@

For the cycling legs of the journey (first and last legs), socio-environmental impacts were estimated on the shifting from car to cycling, using the HEAT for Cycling tool v5.0 [@HEAT] from the World Health Organization, and the [`HEATaaS` R package](https://github.com/HEAT-WHO/HEAT_heatr_api)[^5].
<!-- The HEAT tool provided estimates on the shifting from car to cycling for a short term time horizon (i.e., one year) and the long term (i.e., ten years). -->
We considered two dimensions: *social* — including the physical activity of cyclists, air pollution exposure, and road casualties; *environmental* — including CO~2~eq emissions and other pollutants.

@@

To estimate the car emissions, we used the EMEP/EEA's COPERT software v5 methods and reference values [@COPERT] for a Tier 3 detail level. 
<!-- We used a family-size vehicle, EURO standard, and gasoline or diesel fuel. -->
<!-- All trips were considered to be made under urban conditions and at an average speed of 15 km/h during rush hour periods. -->
<!-- Since the average distance traveled per trip influences the overconsumption and emissions from cold-start engine operation, we estimated energy and emission factors for different ranges of trips at 500-meter intervals. -->
<!-- In addition, we assumed an occupancy rate of 1.6 passengers *per* car [@IMOB]. -->
Regarding PT, we considered the emission factor values reported in the environmental and sustainability reports of the PT operators in the LMA.
<!-- [@Carris2019s; @Metro2019s; @CP2019s; @Transtejo2014]. -->
<!-- Emissions were estimated for the following atmospheric pollutants: CO, PM10, NOx, and VOC; and for the main green house gases: CO~2~, CH~4~, and N~2~O, converted in CO~2~eq. -->
<!-- In particular, for the urban train and tram -- with 100% electric traction -- only CO~2~eq emissions were considered (resulting from the production of electricity, considering a "well-to-tank" approach), since the other pollutants are not emitted locally. -->
The conversion of avoided emissions into avoided welfare loss and respective monetary valuation was based on the EU Guide to Cost-benefit Analysis [@EuropeanCommission2014] and the best up-to-date reference values for the various gases.
<!-- [@EuropeanCommission2014; @bickel2006; @UNITE]. -->
<!-- The social impacts are in avoided premature mortality. This result is finally monetized using the *Statistical Value of Life* for Portugal [@ANSR2021]. -->
<!-- We updated all the monetary reference values of the literature based on the annual inflation rate in Portugal for 2022[^6], and our 10-years estimations assumed a discount rate of 5% and inflation of 3%.  -->

<!-- [^6]: See [Statistics Portugal tool for inflation rate estimates between years](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ipc). -->

@@

```{r summary0}
summary0 = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/HEAT_AML_intermodal_ppp_NEW.Rds"))
# see code/test-code/heat-prep.R#intermodality

summary0 = summary0 |> 
    mutate(Target = paste0(ENMAC*100,"%"),
           Routing = c("safe", "safe", "direct", "direct"),
           TotalTrips = Total,
           Bikebase = Bike,
           Bikepotential = Bike_new,
           Social = Economic - Environmenal_only,
           Social10 = Economic10 - Environmenal10_only) 
```

```{r summary1}
summary1 = summary0 |> 
  select(Target, Routing, TotalTrips, Bikebase, Bikepotential, CO2eq, Economic10) |> 
    arrange(desc(Target))

knitr::kable(summary1 |> mutate(Economic10 = Economic10/1000),
      digits = 0,
      format.args = list(big.mark = " "),
      col.names = c("Target", "Routing", "Total trips",
                    "Baseline Cycling + PT", "Potencial Cycling + PT",
                    "Avoided CO2eq (ton/yr)", "SE Benefits for 10 years (thousand €)"),
      caption = "\\label{summary1}Summary of the cycling potencial of intermodality scenario and its socio-environmental benefits for the cycling legs.",
      format="latex",
      booktabs = TRUE) |> 
  kableExtra::column_spec(column = c(4:7), width = "6em")

@@

<!-- Table \ref{tab:summary21} shows the potential trips by PT mode to replace the second leg of the journey, in combination with cycling. -->
Regarding the PT mode to replace the second leg of the journey, in combination with cycling, trains offer the greatest potential for substitution (88%).
When comparing the existing PT interfaces (Figure \ref{fig:map1}) with the bike routes with highest potential to serve as first and last legs (Figure \ref{fig:map2}) it becomes clear that the Train interfaces are the ones that have the highest potential to attract car-to-PT substituting trips, if their accessibility by bicycle is improved to be safe.

```{r summary21, eval =FALSE}
summary2 = readRDS("load/summary2.Rds")

@@

<!-- Table \ref{tab:summary22} presents the results of the avoided emissions and its monetization for the second leg of the journey, by replacing car trips with potential TP trips. -->
<!-- We estimate a reduction of 8,702 tons of CO2eq emissions by substituting motorized trips relying on fossil fuels and electricity. These estimations used an electricity production life cycle approach. -->
<!-- this table could be split into Social benefits and Environmental benefits -->
Regarding the PT segment, the shift from private car usage would lead to the mitigation of CO~2~ equivalent emissions to 8,500 to 20,800 tons annually, valued in €1.4 million to €3.5 million yearly.

```{r summary22}
summary22 = readRDS("load/TPtable_aml.Rds")
summary22 = summary22 %>%
  mutate(Routing = recode(Percursos, "seguro" = "safe", "direto" = "direct")) |> 
  mutate(Value = money_emissions_eur/1000,
         Target = Meta_ENMAC) |> 
  select(Target, Routing, CO2eq_TP, CO_PT, PM10_PT, NOx_PT, VOC_PT, Value) |> 
  arrange(Target)

# knitr::kable(summary22,
#       digits = c(0,0,0,0,1,0,1,0),
#       format.args = list(big.mark = " "),
#       col.names = c("Target", "Routing", "CO2eq",
#                     "CO", "PM10", "NOx", "VOC", "Value (k€)"),
#       caption = "\\label{summary22}Summary of the avoided emmissions (ton/year) and corresponding monetization (thousand €) by replacing car trips with PT, in the second leg.",
#       format="latex",
#       booktabs = TRUE) |> 
#   kableExtra::column_spec(column = c(3:7), width = "3.5em")
```

@@

```{r summaryall, message=FALSE, warning=FALSE}
rate = ((1+0.03)^10)/((1+0.05)^10) # inflation rate 3% discount rate 5%

summaryall = summary0 |> 
  left_join(summary22) |> 
  select(Target, Routing, CO2eq, CO2eq_TP, Environmenal_only, Environmenal10_only, Social, Social10, Value) |> 
    mutate(Social = Social/1000,
           Social10 = Social10/1000,
         Environmenal_only = Environmenal_only/1000,
         Environmenal10_only = Environmenal10_only/1000) |> 
  mutate(CO2eqT = CO2eq + CO2eq_TP,
         value1 = Environmenal_only + Value + Social, # this is CORRECT
         value10 = Environmenal10_only + Value*10*rate + Social10) |> # this is CORRECT
  select(Target, Routing, CO2eqT, value1, value10) |> 
  arrange(desc(Target))

@@

<!-- FOR OTHER PAPER: separate the Socio impacts from the Environmental ones? It shouldn't be ok to sum them -->
Shifting from car to cycling + PT can reduce annual CO~2~eq emissions emissions by 11,500 to 28,500 tons per year, 
and the 10-year socio-environmental benefits account for €125 million to €325 million, depending on the cycling targets

The environmental impacts represent less than 2% of the socio-environmental benefits (in value) from replacing car trips to bicycle in first-and-last legs.
For the PT segment, we did not estimate the social impacts from substituting car trips, although its health benefits would not be as high as shifting to cycling.<!-- We should estimate the social impacts from replacing Car to PT in the second leg, for the other paper --> <!--Overall there was very little evidence available on mode switch to public transport. -->

The emissions of CO~2~eq that are avoided during both the initial and final journey segments account for about 74% of the emissions avoided during the PT segment. This finding, while expected -- due the zero cycling emissions, should not be overlooked when promoting the PT use.
Improving the safe accessibility to PT interfaces to cyclists and providing bicycle-friendly amenities such as parking facilities can potentially lead to a higher reduction in CO~2~eq emissions, compared to a scenario where individuals shift from car travel to car + PT combination.

Our findings show that cycling *in combination* with PT could replace 10% of current LMA trips, with an additional 6% of PT journeys prone to further substitution, based on conservative assumptions.
